Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-10-10 analysis/paper/paper.tex
casi casi final draft

\title{Asssesment of zonal symmetric and asymmetric components of the Southern
Annular Mode using a novel approach}

@@

\extraauthor{Carolina Vera
}

@@

The Southern Annular Mode (SAM) is the main mode of variability in the
Southern Hemisphere extratropical circulation \citep{rogers1982} in
daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006}
and exerts an important influence in weather conditions such as
temperature and precipitation anomalies and sea ice concentration
\citep{fogt2020}. Its positive phase is traditionally described as
anomalously low pressures over Antarctica surrounded by a ring of
anomalous high pressures in middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, the spatial the
spatial structure of the SAM computed from them contains noticeable
deviations from zonal symmetry, particularly in the Pacific Ocean
region. These zonal asymmetries have not been widely studied, but
previous work suggest that they strongly modulate the regional impacts
of the SAM, going as far as reversing its relationship between
precipitation in South America \citep{silvestri2009}. At the very least,
the fact that the SAM is not entirely zonally symmetric hinders our
ability to reconstruct its historical variability prior to the
availability of dense observations in the Southern Hemisphere
\citep{jones2009}.

At least some of the variability associated with the zonal asymmetries
of the SAM is probably forced by the tropics. In particular, ENSO-like
variability affects the Southern Hemisphere extratopics through the
Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
whose wave train protects strongly onto the zonal anomalies
corresponding to the SAM in the Pacific sector. And although the
relationship between ENSO and SAM is far from simple, tropical
influences on the SAM have been observed
\citep{fan2007, fogt2011, clem2013}. In particular, \citet{fan2007}
computed SAM indices of the Western and the Eastern Hemisphere
separately and found that they were much more correlated if the (linear)
signal of the ENSO was removed. While this relates to temporally
coherent variability of the two hemisphere and not necessarily to zonal
asymmetries in the associated spatial patterns, it is nonetheless
consistent of the ENSO playing a crucial role in zonal asymmetries of
the SAM.

Positive trends in SAM index have been documented by various researchers
using different indices mostly on boreal Summer and Autumn \citep[ and
references therein]{fogt2020}. These trends are thought of driven
primarily by stratospheric ozone depletion and understood in the context
of zonal mean variables. However, it's not clear how or if the
asymmetric component responds to this forcing, or whether its
variability could be masking influencing the observed trends
independently.

Similarly unclear are the specific impacts of the zonally asymmetric
component of the SAM. Positive phase of the SAM is associated with
generally colder temperatures over Antarctica and warmer temperatures at
higher latitudes \citep{jones2019} (and vice versa for negative SAM),
but there are significant deviations from this zonal mean response,
notably in the Antarctic Peninsula and the South Atlantic
\citep{fogt2012}. The SAM signal in precipitation behaves similarly,
although with even greater deviation from zonal symmetry
\citep{lim2016}. The importance of zonal asymmetries of the SAM in these
impacts have been studied in certain regions. For example, the
SAM-precipitation relationship in Southeastern South America and
Southern Brazil can be explained by the PSA-like zonally asymmetric
circulation associated with the SAM \citep{silvestri2009, rosso2018}.
\citet{fan2007} also found that precipitation in East Asia was impacted
by the variability of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal
variability of the asymmetric component of the SAM with the exception of
\citet{fogt2012}. However, their methods based on composites of positive
and negative SAM events leads to some issues, such as spatial patterns
derived from as little as 4 cases and from imbalanced periods (for
example, 5 of the 7 cases in their DJF SAM+ composite are from later
than 1988, whereas all of the 8 years in their DJF SAM- composite are
from earlier than 1988). This is particularly important due to the
inhomogeneities in reanalysis products prior to the satellite era and
the possible change in the asymmetric structure of the SAM
\citep{silvestri2009}.

Moreover, \citet{fogt2012} studied the zonal asymmetric component of the
SAM only in sea level pressure. Zonal asymmetries in the SAM spatial
pattern are fairly barotropic throughout the troposphere, but they
change dramatically in the stratosphere \citep{baldwin2009}.

Our objective is, then, to systematically characterise the zonally
asymmetric and symmetric components of the SAM variability. For each
level, we construct two indices which aim to capture exclusively the
variability of the symmetric and asymmetric component respectively. We
assess their vertical structure and coherence, temporal variability and
trends. We then study the spatial patterns described by the variability
exclusive to each index. Finally, we investigate their relationship with
temperature and precipitation anomalies.

@@

To describe the relationship between the SAM indices and temperature and
precipitation, we use temperature data from NOAA's Merged Land Ocean
Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
which blends land and ocean temperature analysis into a monthly global
grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall
0.5\degree~longitude by 0.5\degree~latitude data from the Global
Precipitation Climatology Centre \citep{schneider2015}.

@@

\begin{figure*}
\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.}\label{fig:method}
\end{figure*}

Our method assumes linearity in the asymmetric component of the SAM.
That is, we assume that zonal symmetries associated with positive SAM
are almost opposite and of the same magnitude to the ones associated
with negative SAM. \citet{fogt2012}'s composites (their Figure 4)
suggest that this might not be entirely valid, although we argue that
much of that apparent non-linearity is due to the heterogenous nature of
the selected years for constructing the composites. Using our data (from
1979 to 2018), seasonal composites of zonal anomalies of 700 hPa
geopotential height for SAM+ (Full SAM index greater than 1 standard
deviation) and SAM- (less than minus 1 standard deviation) show
relatively high pattern correlations for all seasons and are visually
very linear (Figure A9). Therefore, we believe that our method is a
reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are
assuming that the zonal anomalies of the SAM are the same in all seasons
--December to February (DJF), March to May (MAM), June to August (JJA)
and September to November (SON). Geopotential zonal anomalies computed
by projecting the first EOF \emph{of each season} are very similar to
each other (Figure A10) and show pattern correlations between 0.65 (DJF
with JJA) and 0.9 (between MAM and SON). Based on this, we believe that
our initial assumption is not unreasonable.

Finally, we assume that the zonally asymmetric pattern is stationary in
time. \citet{silvestri2009} suggest that this might not be the case
between 1958 and 2004 but the period we analyse is much shorter
(1979-2018) so it's unlikely that we could observe significant changes.
Moreover, zonal asymmetry of the spatial patterns for the two halves of
the period (1979 to 1998 and 1999 to 2018) show no systematic change
(Figure A11).

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
the time, \(\alpha\) and \(\beta\) are the linear regression
coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
From this equation, \(\alpha\) represents the (linear) association of
\(X\) with the variability of the Asymmetric SAM that is not explained
by the variability of the Symmetric SAM; i.e.~it is proportional to the
partial correlation of \(X\) and the Asymmetric SAM, controlling for the
effect of the Symmetric SAM and vice versa for \(\beta\). When
performing a separate regression for each trimester (DJF, MAM, JJA,
SON), we average of the relevant variables seasonally for each year and
trimester before computing the regression.

@@

We computed linear trends by Ordinary Least Squares and the 95\%
confidence interval assuming a t-distribution of the appropriate
residual degrees of freedom.

To compute the amplitude of the zonal waves we computed the Fourier
transform of the spatial field at each latitude circle.

@@

\begin{figure*}
\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa. To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \cite{sheather1991}. Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
\end{figure*}

@@

The stratospheric Symmetric SAM varies strongly with a two-year period,
which can be seen by spectral analysis (Figure A3). This might suggests
a link between stratospheric SAM variability and the Quasi-Biennial
Oscillation \citep{baldwin2001b}. There is a local peak at 2 years in
the periodogram of the tropospheric Symmetric SAM also, although it's
not statistically significant. In the troposphere the most significant
peak of variability is found in the Asymmetric index at around 3.6
months.

@@

\begin{figure}
\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.}\label{fig:cor-lev}
\end{figure}

\begin{figure*}
\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.}\label{fig:cross-correlation}
\end{figure*}

@@

\begin{figure*}
\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
\end{figure*}

The linear trends for each of the indices (Full SAM, Symmetric SAM and
Asymmetric SAM) were evaluated for the complete period 1979-2018 at each
level (Figure \ref{fig:trends}) for the whole year and separated by
trimesters. The Full SAM index presents a statistically significant
trend (panel a.1) that extends throughout the troposphere up to about 50
hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest
of column a) indicate that positive trends are present in Autumn and
particularly in Summer, where the 100 hPa maximum is much more defined.
In Winter and Spring, we detect no statistically significant trend. This
is consistent with the results of previous studies, which find clear
positive trends in Summer, weaker in Autumn and no trends in the other
seasons \citep[e.g.][ and references therein]{fogt2020} using indices of
the SAM based on surface or near-surface circulation.

@@

\begin{figure}
\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018. Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
\end{figure}

@@

\begin{figure*}
\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices. Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index. }\label{fig:2d-regr}
\end{figure*}

@@

\begin{figure*}
\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
\end{figure*}

@@

We can see that the amplitude and latitudinal distribution zonal waves
in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric
SAM on the other correspond almost exactly to the amplitude and
latitudinal distribution in the Full SAM. This confirms the correct
decomposition of the SAM in its symmetric and asymmetric components.

Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent
that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which
--as mentioned before-- is larger in the Western Hemisphere than in the
Eastern Hemisphere.

\begin{figure}
\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM). (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979–2018 period.}\label{fig:vertical-regression}
\end{figure}

@@

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very
similar to the Pacific-South American Pattern \citep{mo1987, kidson1988}
which is a teleconnection pattern associated with the ENSO
\citep{karoly1989}. Indeed, \citet{fogt2011} showed that there is a
significant relationship between the SAM and the ENSO. The correlation
between the Full SAM and the ENSO as measured by the Oceanic Niño Index
\citep{bamston1997} (ONI) is -0.16 (p-value =
\ensuremath{2.8\times 10^{-4}}). Consistent with \citet{fan2007}, we
show that this relationship is captured entirely by the Asymmetric SAM,
as this index has a partial correlation of -0.26 (p-value =
\ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect
of the Symmetric SAM, whereas the Symmetric SAM's partial correlation
with the ONI is essentially null (0.019; p-value = 0.67). We performed
the same analysis using the Multivariate ENSO Index \citep{wolter2011}
and the Southern Oscillation Index \citep{ropelewski1987} to conclude
that these results do not depend on the ENSO index used.

\subsection{Impacts}

@@

\begin{figure*}
\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period. Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate. Gray areas in Antarctica are areas with have more than 15\% of missing data.}\label{fig:regr-air-season}
\end{figure*}

@@

The pattern of negative anomalies in the pole surrounded by positive
anomalies roughly seen in all seasons --although with varying intensity
and small-scale details-- translates to enhanced meridional temperature
gradient maximised in the zero line, which is consistent with the
intensification and poleward migration of the westerlies commonly linked
to the SAM through thermal wind balance. It's then not surprising to see
it more clearly in association with the Symmetric SAM (at least in
Summer and Spring).

@@

\begin{figure*}
\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period. Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
\end{figure*}

@@

\begin{figure*}
\includegraphics{pp-regr-america-1} \caption{Same as Figure \ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
\end{figure*}

In South America (Figure \ref{fig:pp-regr-america}), the annual-level
regression shows that positive SAM is associated with statistically
significant precipitation decrease in Southeastern South America (SESA)
and Southern Chile and non-significant increase in South Brazil, near
the South Atlantic Convergence Zone (SACZ) (panel a.1). Panels b.1 and
c.1 show a remarkably clean separation between the Asymmetric SAM
--associated with the Southeastern South American and Southern Brazilian
signals-- and the Symmetric SAM --associated with the signal in Southern
Chile.

>>>>

\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach}

@@

\extraauthor{Carolina Vera
}
\extraaffil{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina}

@@

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation \citep{rogers1982} in daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006} and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration \citep[e.g.][]{fogt2020}. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with these indices (either by regressino or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region. These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep[\citet{fogt2012}, \citet{rosso2018}]{fan2007}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.

At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics. In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989}, whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector. And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}. In particular, \citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed. While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.

Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}. These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}. However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.

Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM. Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}. The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}. The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions. For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}. \citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of \citet{fogt2012}. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM \citep{silvestri2009}. Moreover, \citet{fogt2012} studied the zonal asymmetric component of the SAM only in sea level pressure. Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere \citep{baldwin2009}.

Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability. For each level, we construct two indices which aim to capture exclusively the variability of the symmetric and asymmetric component respectively. We assess their vertical structure and coherence, temporal variability and trends. We then study the spatial patterns described by the variability exclusive to each index. Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.

\#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro. Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos. Por ej. para la evaluación de los impactos solamente se está considerando el ínidice troposférico.

@@

To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017}. The rainfall product uses station-based records, and thus it only has continental coverage.

@@

\begin{figure*}
\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.}\label{fig:method}
\end{figure*}

Our method assumes linearity in the asymmetric component of the SAM. That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM. \citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4). Therefore, we believe that our method is a reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year. Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA). Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.

Finally, we assume that the zonally asymmetric pattern is stationary in time. \citet{silvestri2009} suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes. Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms. From this equation, \(\alpha\) represents the (linear) association of \(X\) with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.~it is proportional to the partial correlation of \(X\) and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for \(\beta\). When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.

@@

We computed linear trends by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom. To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.

We computed density probability estimates using a gaussian kernel of optimal bindwidth according to \citet{sheather1991}.

@@

\begin{figure*}
\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa. To the right, probability density estimate of each index. Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
\end{figure*}

@@

The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2). This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}. There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant. In the troposphere the most significant peak of variability is found in the Asymmetric index at around 3.6 months.

@@

\begin{figure}
\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
\end{figure}

\begin{figure*}
\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
\end{figure*}

@@

\begin{figure*}
\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
\end{figure*}

The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level (Figure \ref{fig:trends}) for the whole year and separated by trimesters. The Full SAM index presents a statistically significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest of column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined. In Winter and Spring, we detect no statistically significant trend. This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.

@@

\begin{figure}
\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018. Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
\end{figure}

@@

\begin{figure*}
\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices. Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index. }\label{fig:2d-regr}
\end{figure*}

@@

\begin{figure*}
\includegraphics{wave-amplitude-1} \caption[Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean]{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
\end{figure*}

@@

We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM. This confirms the correct decomposition of the SAM in its symmetric and asymmetric components. Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.

\begin{figure}
\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM). (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
\end{figure}

@@

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}. Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO. The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}). Consistent with \citet{fan2007}, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67). We performed the same analysis using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987} to conclude that these results do not depend on the ENSO index used.

\subsection{Surface Impacts}

@@

\begin{figure*}
\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period. Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate. Gray areas in Antarctica are areas with have more than 15\% of missing data.}\label{fig:regr-air-season}
\end{figure*}

@@

The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance. It's then not surprising to see it more clearly in association with the Symmetric SAM (at least in Summer and Spring).

@@

\begin{figure*}
\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period. Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
\end{figure*}

@@

\begin{figure*}
\includegraphics{pp-regr-america-1} \caption[Same as Figure \\ref{fig:pp-regr-oceania} but for South America]{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
\end{figure*}

In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1). Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM -- associated with the Southeastern South American and Southern Brazilian signals -- and the Symmetric SAM -- associated with the signal in Southern Chile.
